---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---



## Loading and preprocessing the data
```{r}
mydat <- read.csv("activity.csv")
# cleanup date
mydat$date <- as.Date(as.character(mydat$date))

#clean NA's...set to zero
newdat <-mydat[!is.na(mydat$steps),]

#use plyr to get sum, mean and median number of steps each day
library(plyr)
datsum<-ddply(newdat,"date",summarize,totsteps=sum(steps))
datmean<-ddply(newdat,"date",summarize,meansteps=mean(steps))
datmedian<-ddply(newdat,"date",summarize,medsteps=median(steps))

```


## Histogram of the total number of steps taken each day
```{r}

  hist(datsum$totsteps)
# calc mean and median numer of total steps per day
meansteps <-mean(datsum$totsteps)
mediansteps <- median(datsum$totsteps)
```

## What is the mean total number of steps taken per day?
The mean total number of steps is **`r meansteps`**

The median total number of steps is **`r mediansteps`**


## What is the average daily activity pattern?
```{r}
# calculate mean #steps for each 5-minute interval

minuteData <- ddply(newdat,"interval",summarize,minavg=mean(steps))
plot(minuteData$interval,minuteData$minavg,type='l',xlab='interval',ylab='Average steps')

# determine which interval contains the maximum number of steps
maxInterval <- minuteData$interval[which(minuteData$minavg == max(minuteData$minavg))]
```

The interval with the maximum number of steps is **`r maxInterval`**

-------

## Imputing missing values
```{r}
# calculate number of missing samples
misssamp <- sum(is.na(mydat$steps))

```
There are  **`r misssamp`** missing samples

## Are there differences in activity patterns between weekdays and weekends?
